---
title: Simulate data and fit a 2-species static (aka single-season) occupancy model à la Rota et al. (2016)
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 600, cache = TRUE, warning = FALSE, message = FALSE)
```

## Setting the scene

We consider a two-species static occupancy model à la [Rota et al. (2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12587). We simulate data from this model, and fit the model to these data using `Unmarked`.

Ignoring the site index, we use the following notation for the occupancy probabilities:

* $\psi_{11}$ is the prob. that species 1 and species 2 are both present;  
* $\psi_{10}$ is the prob. that species 1 is present and species 2 is absent;
* $\psi_{01}$ is the prob. that species 1 is absent and species 2 is present;
* $\psi_{00}$ is the prob. that species 1 and species 2 are both absent,
with avec $\psi_{11} + \psi_{10} + \psi_{01} + \psi_{00} = 1.$

The marginal probabilities of occupancy are:

* $\Pr(z_1 = 1) = \Pr(\mbox{species 1 is present}) = \psi_{10} + \psi_{11}$
* $\Pr(z_2 = 1) = \Pr(\mbox{species 2 is present}) = \psi_{01} + \psi_{11}$
* $\Pr(z_1 = 0) = \Pr(\mbox{species 1 is absent}) = \psi_{01} + \psi_{00}$
* $\Pr(z_2 = 0) = \Pr(\mbox{species 2 is absent}) = \psi_{10} + \psi_{00}$

And the conditional probabilities (reminder: $\Pr(\mbox{A|B}) = \Pr(\mbox{A and B})/\Pr(\mbox{B})$):

* $\Pr(z_1 = 1 | z_2 = 0) = \psi_{10} / (\psi_{10} + \psi_{00}) = \Pr(\mbox{species 1 is present given species 2 is absent});$
* $\Pr(z_1 = 1 | z_2 = 1) = \psi_{11} / (\psi_{11} + \psi_{01}) = \Pr(\mbox{species 1 is present given species 2 is present});$
* $\Pr(z_2 = 1 | z_1 = 0) = \psi_{01} / (\psi_{01} + \psi_{00}) = \Pr(\mbox{species 2 is present given species 1 is absent});$
* $\Pr(z_2 = 1 | z_1 = 1) = \psi_{11} / (\psi_{11} + \psi_{10}) = \Pr(\mbox{species 2 is present given species 1 is present}).$

## Data simulation

We will use the package `mipfb` to simulate occupancy state as a multivariate Bernoulli random variable; more about the multivariate Bernoulli can be found in [Dai et al. (2013)](https://arxiv.org/pdf/1206.1874.pdf):
```{r}
library(mipfp)
```

For reproducibility, we set the seed:
```{r}
set.seed(2020) 
```

Choose the number of species, the number of sites, and the number of visits: 
```{r}
S <- 2 # nb species 
N <- 500 # nb sites
J <- 5 # nb visits
```

Let's consider a scenario in which species 2 avoids species 1 while species 1 does not care about species 2 and its presence or absence. To specify this scenario, we will work out the conditional probabilities with, for example:

* $\Pr(z_2 = 1 | z_1 = 0) = 0.6$, species 2 is present with high probability whenever species 1 is absent
* $\Pr(z_2 = 1 | z_1 = 1) = 0.1$, species 2 avoids species 1 when it is present
* $\Pr(z_1 = 1 | z_2 = 0) = \Pr(z_1 = 1 | z_2 = 1) = 0.4$, species 1 does not care about presence/absence of species 2

Now we need to go back to the probabilities of occupancy. Let $x = \psi_{01}$, $y = \psi_{10}$ et $z = \psi_{11}$ soit $1 - x - y - z = \psi_{00}$, then we have a system of 3 equations with 3 unknowns:

$$0.6 = x / (x + 1 - x - y - z) \Leftrightarrow x + 0.6y + 0.6z = 0.6 \\
0.1 = z / (z + y) \Leftrightarrow -0.1y + 0.9z = 0 \\
0.4 = y / (y + 1 - x - y - z) \Leftrightarrow 0.4x + y + 0.4z = 0.4$$

which can be solved with the [Mathematica online solver](https://www.wolframalpha.com/input/?i=solve%7Bx%2B0.6y%2B0.6z%3D%3D0.6%2C-0.1y%2B0.9z%3D%3D0%2C0.4x%2By%2B0.4z%3D%3D0.4%7D): 
```{r}
psi01 <- 81/175
psi10 <- 36/175
psi11 <- 4/175
psi00 <- 1 - (psi01 + psi10 + psi11) # 54/175
```

We then obtain the marginal occupancy probabilities:
```{r}
psiS1 <- psi10 + psi11
psiS2 <- psi01 + psi11
```

Now we're ready to simulate data from a multivariate Bernoulli (check out `?RMultBinary` and `?ObtainMultBinaryDist`).

First, we calculate the odds ratios:
```{r}
or <- matrix(c(1, (psiS1*(1-psiS2))/(psiS2*(1-psiS1)), 
               (psiS2*(1-psiS1))/(psiS1*(1-psiS2)), 1), nrow = 2, ncol = 2, byrow = TRUE)
rownames(or) <- colnames(or) <- c("sp1", "sp2")
```

Then the marginal probabilities:
```{r}
marg.probs <- c(psiS1, psiS2)
```

And we estimate the joint probability:
```{r}
p.joint <- ObtainMultBinaryDist(odds = or, marg.probs = marg.probs)
```

At last, we generate $N$ random samples from a bivariate Bernoulli (2 species) with relevant parameters
```{r}
z <- RMultBinary(n = N, mult.bin.dist = p.joint)$binary.sequences 
```

Now we add on top the observation. First, we fix the detection probability for each species:
```{r}
ps <- c(0.5,0.9)
```

Then we generate the detection and non-detections for each species, which we store in a list: 
```{r}
y <- list()
for (i in 1:S){
  y[[i]] <- matrix(NA,N,J)
  for (j in 1:N){
    for (k in 1:J){
      y[[i]][j,k] <- rbinom(1,1,z[j,i]*ps[i])
    }
  }
}
names(y) <- c('sp1','sp2')
```

## Model fitting

Now let us fit a 2-species static occupancy model to the data we have simulated. We need to load the package `unmarked`:
```{r}
library(unmarked)
```

We format the data as required:
```{r}
data <- unmarkedFrameOccuMulti(y=y)
```

Let's have a look to the data:
```{r}
summary(data)
```

And in particular the detections and non-detections:
```{r}
plot(data)
```

Now we specify the effects we would like to consider on the occupancy and detection probabilities. The thing is that the function `occuMulti` doesn't work directly on the occupancy probabilities but on the so-called natural parameters (in that specific order): 

* $f_1 = \log(\psi_{10}/\psi_{00})$;
* $f_2 = \log(\psi_{01}/\psi_{00})$;
* $f_{12} = \log(\psi_{00}\psi_{11} / \psi_{10}\psi_{01})$, 

that is:

* $\psi_{11} = \exp(f_1+f_2+f_{12})/\mbox{den}$;
* $\psi_{10} = \exp(f_1)/\mbox{den}$;
* $\psi_{01} = \exp(f_2)/\mbox{den}$,
where $\mbox{den} = 1+\exp(f_1)+\exp(f_2)+\exp(f_1+f_2+f_{12})$:
```{r}
occFormulas <- c('~1','~1','~1') 
```

To specify the effects on detection, there is no difficulty:
```{r}
detFormulas <- c('~1','~1')
```

We fit a model with constant natural parameters and constant detection probabilities
```{r}
fit <- occuMulti(detFormulas,occFormulas,data)
```

Display the result:
```{r}
fit
```

Get the natural parameter and detection estimates:
```{r}
mle <- fit@opt$par
names(mle) <- c('f1','f2','f12','lp1','lp2')
```

Get the occupancy estimates:
```{r}
den <- 1 + exp(mle['f1'])+exp(mle['f2'])+exp(mle['f1']+mle['f2']+mle['f12'])
psi11hat <- exp(mle['f1']+mle['f2']+mle['f12'])/den
psi10hat <- exp(mle['f1'])/den
psi01hat <- exp(mle['f2'])/den
```

I do it by hand to understand how `unmarked` works. The easy way is to use `predict(fit,'state')`.

Get the detection estimates:
```{r}
p1hat <- plogis(mle['lp1'])
p2hat <- plogis(mle['lp2'])
```

Again I do it by hand, but `unmarked` can do it for you with `predict(fit,'det')`.

Now compare the parameters we used to simulate the data (left column) to the parameter estimates (right column)
```{r}
res <- data.frame(real = c(psiS1,
                           psiS2,
                           psi01,
                           psi10,
                           psi11,
                           ps[1],
                           ps[2]),
                  estim = c(psi10hat+psi11hat,
                            psi01hat+psi11hat,
                            psi01hat,
                            psi10hat,
                            psi11hat,
                            p1hat,
                            p2hat))
rownames(res) <- c('marginal_occ1','marginal_occ2','psi01','psi10','psi11','det1','det2')
res
```

If you just want to get the parameter estimates directly:
```{r}
# detection
predict(fit,'det',species=1)[1,]
predict(fit,'det',species=2)[1,]

# marginal occupancy
predict(fit,'state',species=1)[1,]
predict(fit,'state',species=2)[1,]

# conditional occupancy
predict(fit,'state',species=1,cond='sp2')[1,] # species 1 | species 2 present
predict(fit,'state',species=1,cond='-sp2')[1,] # species 1 | species 2 absent
predict(fit,'state',species=2,cond='sp1')[1,] # species 2 | species 1 present
predict(fit,'state',species=2,cond='-sp1')[1,] # species 2 | species 1 absent
```

## R version used

```{r}
sessionInfo()
```

